\documentclass[11pt]{article}
\usepackage{amsmath,amssymb,amsthm,booktabs,array,longtable,geometry,hyperref,xcolor}
\geometry{margin=1in}

\newtheorem{theorem}{Theorem}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}

\definecolor{corrected}{rgb}{0.0,0.5,0.1}
\definecolor{removed}{rgb}{0.7,0.1,0.1}

\title{Two-Layer Corrections for the Monogate Research Record\\
Systematic Relabeling of Pre-Audit Node Counts}
\author{Monogate Research}
\date{2026-04-21}

\begin{document}
\maketitle

\begin{abstract}
The operator primitiveness audit (Operator\_Primitiveness\_Audit.tex) and
the subsequent decision paper (23op\_Layer\_Status\_Decision.tex) established
that seven operators (EEM, EED, EES, LLA, LLS, LLD, EEA) are not all genuine
F16 primitives.
This document serves as the authoritative correction record for all prior
papers and exploration files.
It identifies every location where (1) a 23-op extended count was presented
as if it were an F16 ground-truth count, (2) node counts were incorrect
under the F16-only model, or (3) savings claims conflated genuine F16 savings
with notation-only savings.
The corrections are organized by file.
After this document, the research record is consistent with the two-layer
accounting policy established in 23op\_Layer\_Status\_Decision.tex.
\end{abstract}

\section{Correction Policy Reminder}

\begin{definition}[Two-layer accounting, restated]
\textbf{Layer 1 (F16 ground-truth)}: counts in terms of the 16 F16 operators only.
Extended operators (EEM, EED, EES, LLA, LLS, LLD, EEA) each cost their F16-decomposition
price (3n, 3n, 4n general, 3n, 3n, 4n, 4n respectively), not 1n.

\textbf{Layer 2 (23-op extended)}: extended operators count as 1n each.
Valid only when labeled explicitly as ``23-op count'' or ``extended-primitive count.''
\end{definition}

The canonical correct F16 counts for commonly used expressions are in
F16\_Only\_High\_Impact\_Audit.tex.
The quantum/physics extended-primitive counts are in
Quantum\_Costs\_Final\_23.tex (Layer 2, now correctly labeled).

\section{File-by-File Corrections}

\subsection{SuperBEST\_v6\_With\_New\_Primitives.tex}

\textbf{Status:} All node count tables in this file use 23-op primitive counts (1n per extended operator)
without explicit labeling.

\textbf{Required correction:} Add the following note to the abstract and each table caption:
\begin{quote}
\textit{Node counts in this table use the 23-operator extended-primitive model
(Layer 2). Each of the operators LLS, EEA, EEM, EED, LLA, LLS, LLD is counted
as 1 node. Corresponding F16 ground-truth counts (Layer 1) are larger by
0--3 nodes per extended operator used; see F16\_Only\_High\_Impact\_Audit.tex.}
\end{quote}

\textbf{Specific corrections:}
\begin{center}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{lccl}
\toprule
Expression & v6 count & F16 correct & Note \\
\midrule
$\ln(x/y)$ (LLS) & 1n & 3n & LLS = div+ln in F16 \\
$e^{x+y}$ (EEM) & 1n & 3n & EEM = add+exp in F16 \\
$e^{x-y}$ (EED) & 1n & 3n & EED = sub+exp in F16 \\
$e^x + e^y$ (EEA) & 1n & 4n & EEA: no F16 shortcut \\
Softplus ($2n$ via EEA+ln) & 2n & 2n & Correct! F16 also 2n via EML$(x,1/e)$+ln \\
KL divergence term & 3n (LLS) & 5n & LLS route: div+ln+mul = 5n \\
\bottomrule
\end{tabular}
\end{center}

\subsection{SuperBEST\_v7\_Full\_23\_Audit.tex}

\textbf{Status:} The file is an audit of the 23-op extended model.
The title and abstract correctly refer to ``23 operators'' but do not include
the Layer 2 label consistently in tables.

\textbf{Required correction:} Retitle each table as
``23-op Extended-Primitive Counts (Layer 2)'' and add a footer to each:
\begin{quote}
\textit{$\dagger$ F16 ground-truth counts (Layer 1) are available in
F16\_Only\_High\_Impact\_Audit.tex and are larger by up to 3n per operator used.}
\end{quote}

\textbf{Specific node count corrections (F16 ground-truth for reference):}
\begin{center}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{lccl}
\toprule
Expression & v7 (23-op) & F16 Layer 1 & Category \\
\midrule
$\ln(xy)$ (LLA) & 1n & 3n & A: mul+ln \\
$\log_y(x)$ (LLD) & 1n & 4n & C: notation-only \\
$e^{x+y}$ (EEM) & 1n & 3n & A: add+exp \\
$e^{x-y}$ (EED) & 1n & 3n & A: sub+exp \\
Quantum rel.\ entropy & 3n (LLS) & 5n & LLS is 3n F16; +mul = 5n \\
Partition fn.\ pair & 7n (EEA) & 10n & EEA is 4n F16; 7+3=10n \\
LSE & 2n (EEA+ln) & 5n & EEA+ln F16 = 4+1 = 5n \\
Spectral decay & 3n (EED) & 5n & EED = sub+exp = 3n; +others \\
\bottomrule
\end{tabular}
\end{center}

\subsection{SuperBEST\_v8\_Quantum\_Physics\_Integration.tex}

\textbf{Status:} This is the most extensive pre-audit paper.
It reports aggregate savings of $\sim40\%$ over a representative system using
the 23-op extended model. The aggregate is correct \emph{within} the 23-op model
but overstates genuine F16 savings.

\textbf{Required correction:}

(1) Abstract: replace ``\textit{40\% savings over the 16-operator baseline}''
with ``\textit{40\% savings in the 23-op extended-primitive model (Layer 2);
genuine F16-only savings are approximately 8--15\% for the same expression set.}''

(2) Table header: add column label ``(23-op count, Layer 2)''.

(3) Summary sentence: ``230n (23-op) for representative system'' is correct
for Layer 2 but the Layer 1 (F16) total is approximately 300--330n for the
same system (the exact count depends on which EEA and LLS uses are genuine
Category A saves vs.\ notation-only Category C).

\textbf{Key per-expression corrections:}
\begin{center}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{lcccc}
\toprule
Expression & v8 (23-op) & F16 (Layer 1) & Difference & Type \\
\midrule
Quantum rel.\ entropy & 3n & 5n & $+2n$ & C: LLS notation \\
Partition fn.\ (per pair) & 7n & 10n & $+3n$ & C: EEA notation \\
LSE & 2n & 5n & $+3n$ & C: EEA notation \\
Von Neumann entropy & 5n & 8n & $+3n$ & C: LLS+EEA notation \\
Bures distance & 3n & 4n & $+1n$ & A: genuine F16 save \\
Boltzmann ratio & 5n & 6n & $+1n$ & C: EED notation (4n genuine) \\
Mayer $f$-function & 3n & 3n & $0$ & B: genuine F16 via EML$_\mathrm{neg}$ \\
Softplus & 2n & 2n & $0$ & B: genuine F16 via EML$(x,1/e)$ \\
Kraus $\sqrt{1-e^{-\gamma t}}$ & 4n & 4n & $0$ & B: genuine F16 \\
Hydrogen radial & 3n & 3n & $0$ & B: genuine F16 \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Quantum\_Costs\_Final\_23.tex}

\textbf{Status:} Written as a 23-op extended-model document.
The title says ``Final 23'' which partially signals the model but the
abstract does not include the Layer 2 label.

\textbf{Required correction:} Add to abstract:
\begin{quote}
\textit{All node counts in this document are 23-op extended-primitive counts (Layer 2).
Genuine F16 savings (Layer 1) are identified in F16\_Only\_High\_Impact\_Audit.tex.}
\end{quote}

The following counts from Quantum\_Costs\_Final\_23.tex were found to be
\emph{overstated} relative to F16 in the audit:
\begin{center}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{lccc}
\toprule
Item & Claimed savings (23-op) & Genuine F16 savings & Type \\
\midrule
Quantum rel.\ entropy & $6n \to 3n$ (save $3n$) & $6n \to 5n$ (save $1n$) & C \\
Partition fn.\ pair & $10n \to 7n$ (save $3n$) & $10n \to 10n$ (save $0n$) & C \\
Lindblad dissipator & $12n \to 9n$ (save $3n$) & $12n \to 11n$ (save $1n$) & C \\
Kraus factor & $7n \to 4n$ (save $3n$) & $6n \to 4n$ (save $2n$) & B \\
\bottomrule
\end{tabular}
\end{center}
These overclaims are notation-only (Category C) except Kraus which is genuine.

\subsection{SpecialFunctions\_Physics\_Final\_Audit.tex}

\textbf{Status:} Mostly correct. The special function $\infty$-cost results
(erf, Bessel, Airy, lgamma, Fresnel, $\zeta$) are unconditionally correct and
do not involve extended operators.

\textbf{No correction required for} the $N(f) = \infty$ results.

\textbf{Correction required for:}
\begin{center}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{lccc}
\toprule
Expression & Claimed cost & F16 correct & Note \\
\midrule
Boltzmann ratio $e^{-\beta(E_j - E_i)}$ & $5n$ (via EED) & $6n$ F16 & EED notation (+1n); genuine save is 4n \\
Mayer $f$: $e^{-\beta u}-1$ & $3n$ (via EES) & $3n$ F16 genuine & B: EML$_\mathrm{neg}$ achieves same \\
Saddle-point integrand & $7n \to 6n$ & $7n \to 7n$ & EED used; no genuine F16 gain \\
Hydrogen radial & $4n \to 3n$ & $4n \to 3n$ & B: genuine EPL save \\
\bottomrule
\end{tabular}
\end{center}

\subsection{New\_Minimal\_Identities\_23.tex}

\textbf{Status:} Mixed. Several identities use 23-op counts correctly;
others are genuine F16 results.

\textbf{Confirmed correct (genuine F16):}
\begin{itemize}
  \item Softplus $= 2n$: F16 genuine via EML$(x, 1/e)$ + ln (Category B)
  \item $x^c = 1n$: F16 genuine via EPL$(c, x)$
  \item $e^{cx} = 1n$: F16 genuine via EPL with constant
  \item $\ln(xy) = 3n$: F16 correct (LLA = mul+ln)
\end{itemize}

\textbf{Requires 23-op label:}
\begin{itemize}
  \item LSE($x,y$) $= 2n$: only under 23-op model (EEA+ln = 1n+1n); F16 = 5n
  \item KL term $= 3n$: only under 23-op (LLS+mul = 1n+2n); F16 = 5n
  \item Telescoping log-ratio $= 1n$: only under 23-op (LLS); F16 = 3n per term
  \item Two-arg softmax $= 3n$: only under 23-op; F16 = 6n
\end{itemize}

\subsection{preprint.tex (main arXiv paper)}

\textbf{Status:} The main preprint does not use 23-op extended counts in its
core claims. The SuperBEST core table (15n/79.5\%) is F16 ground-truth and
is correct as stated.

\textbf{No correction required} to the core table or the 79.5\% headline figure.

\textbf{Minor clarification required:} The phrase
``extended by softplus for numerical continuity'' (line 630) is fine as stated
(softplus is 2n in both models) but should note:
``softplus $\ln(1+e^x)$ costs $2n$ in both F16 and 23-op models via
EML$(x, 1/e) + \ln(\cdot)$.''

\section{What Remains Correct Without Correction}

\begin{enumerate}
  \item \textbf{SuperBEST core table} (v5.2): 15n/79.5\% positive, 22n/69.9\% general.
    All F16 ground-truth. Unaffected.
  \item \textbf{sqrt = 1n} via EPL$(1/2, x)$. Correct.
  \item \textbf{$N(\sin) = N(\cos) = N(\tan) = \infty$}. Proved. Correct.
  \item \textbf{$N(\mathrm{erf}) = N(J_0) = N(\mathrm{Ai}) = \infty$}. Proved. Correct.
  \item \textbf{Mayer $f$-function $= 3n$ in F16}. Genuine Category B. Correct.
  \item \textbf{Softplus $= 2n$ in F16}. Genuine Category B. Correct.
  \item \textbf{Boltzmann ratio: 4n genuine F16 saving} ($10n \to 6n$). Correct.
  \item \textbf{Kraus factor $= 4n$ in F16}. Genuine Category B. Correct.
  \item \textbf{Hydrogen radial $= 3n$ in F16}. Genuine. Correct.
  \item \textbf{tan(1) $\notin \varepsilon(\mathbb{R})$}. Proved. Correct.
  \item \textbf{Three-tier ELC complement classification}. Correct.
  \item \textbf{CONJ\_NO\_OP\_24 (now Theorem)}: exactly 23 operators. Correct.
\end{enumerate}

\section{Corrected Headline Numbers}

\begin{center}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{lccc}
\toprule
Claim & Pre-audit & F16-correct (Layer 1) & 23-op (Layer 2) \\
\midrule
SuperBEST positive total & 15n & 15n & 15n \\
SuperBEST savings & 79.5\% & 79.5\% & 79.5\% \\
Quantum rel.\ entropy (per term) & 3n & 5n & 3n \\
Partition fn.\ pair & 7n & 10n & 7n \\
LSE($x,y$) & 2n & 5n & 2n \\
Softplus & 2n & 2n & 2n \\
Mayer $f$-function & 3n & 3n & 3n \\
Boltzmann ratio (full) & 5n & 6n & 5n \\
Aggregate quantum system & $\sim 230n$ & $\sim 300$--$330n$ & $\sim 230n$ \\
Aggregate savings & 40\% & 8--15\% genuine & 40\% (23-op) \\
\bottomrule
\end{tabular}
\end{center}

\textbf{Note}: The 40\% savings in the 23-op model is real \emph{for implementations
that provide the 7 extended operators as atomically fused primitives} (e.g., CUDA kernels,
hardware-fused DEML/EEA/LLS operations). The 8--15\% genuine F16 savings applies to
software using standard exp/ln library calls.

\section{Publication Status Recommendation}

\begin{enumerate}
  \item \textbf{arXiv preprint (preprint.tex):} No corrections needed. The core
    F16 results (15n/79.5\%) are sound. Proceed to submission as-is for the
    core complexity results. Addenda for the quantum catalog should be filed
    separately as Layer 2 extended-model results.

  \item \textbf{Extended catalog papers (v6, v7, v8):} Relabel all tables
    as 23-op Layer 2 before any public filing or linking from arXiv.

  \item \textbf{New quantum results paper:} Can be written from scratch using
    F16\_Only\_High\_Impact\_Audit.tex (Layer 1) as the ground-truth,
    with a separate section for Layer 2 results clearly labeled.
    The genuine F16 savings (Mayer $f$, softplus, Boltzmann, Kraus) are real
    and publishable as F16 results.
\end{enumerate}

\end{document}
