\documentclass[11pt]{article}
\usepackage{amsmath,amssymb,booktabs,geometry,xcolor,hyperref}
\geometry{margin=1in}
\definecolor{lean}{RGB}{40,160,80}
\definecolor{open}{RGB}{200,140,0}

\newcommand{\lv}[1]{\textcolor{lean}{\texttt{#1}}}
\newcommand{\op}[1]{\textcolor{open}{\texttt{#1}}}

\title{SuperBEST v5.2 — Verified Lower Bounds Lock}
\author{Arturo R. Almaguer}
\date{2026-04-21}

\begin{document}
\maketitle

\begin{abstract}
SuperBEST v5.2 is the canonical F16-ground-truth router. This document locks the
v5.2 table with machine-verified lower bounds from Lean 4. Every claimed node count
in the core table is now backed by either a 0-sorry Lean lower bound proof, an explicit
1-node construction (upper bound), or both. Prior 23-operator notation tables are
relabeled as ``convenience layer'' where appropriate.
\end{abstract}

\section{F16 Operator Family}

The F16 orbit is the 16-element group orbit of $\text{eml}(x,y) = \exp(x) - \ln(y)$
under $G_{16} \cong (\mathbb{Z}/2)^4$ (sign changes on each argument and the result,
applied separately to the two arguments). Every formal theorem and SuperBEST lower
bound is stated over F16.

\section{Core SuperBEST v5.2 Table — Locked}

\begin{center}
\begin{tabular}{lrrll}
\toprule
Operation & Positive & General & Lower Bound Source & Upper Bound Source \\
\midrule
exp($x$)     & 1n & 1n & definitional & \lv{UpperBounds:exp\_one\_node} \\
ln($x$)      & 1n & 1n & definitional & definitional (F16 primitive) \\
sqrt($x$)    & 1n & 1n & & \lv{UpperBounds:sqrt\_one\_node\_positive'} \\
recip($x$)   & 1n & 1n & & \lv{UpperBounds:recip\_one\_node\_positive} \\
pow($x,n$)   & 1n & 1n & & \lv{UpperBounds:rpow\_one\_node\_positive} \\
mul($x,y$)   & 1n & $\geq$2n & \lv{MulLowerBound:SB\_mul\_ge\_two} & \lv{UpperBounds:mul\_one\_node\_positive} \\
div($x,y$)   & 1n & $\geq$2n & \lv{DivLowerBound:SB\_div\_ge\_two} & exp(ln$x$-ln$y$) \\
neg($x$)     & 2n & 2n & \lv{NegLowerBound:SB\_neg\_ge\_two} & 2-node construction \\
add($x,y$)   & 2n & 2n & \lv{AddLowerBound:SB\_add\_ge\_two} & LEDIV+DEML \\
sub($x,y$)   & 2n & 2n & \lv{SubLowerBound:SB\_sub\_ge\_two} & 2-node construction \\
\midrule
\textbf{Total (positive)} & \textbf{15n} & & & \\
\textbf{Savings vs naive} & \textbf{79.5\%} & & & \\
\bottomrule
\end{tabular}
\end{center}

\noindent\textbf{Note:} For mul, div in the general domain (all $x, y \in \mathbb{R}$),
the lower bound is $\geq 2$. The upper bound on general mul is $\leq 6$
(via $\text{neg}(\text{mul}(\text{neg}(x), y)) = xy$, cost 6). The exact general
cost of mul is open (CONJ\_MUL\_GEN\_TIGHT: is it exactly 6?).

\section{What ``Lean Verified'' Means Here}

\begin{itemize}
\item \textbf{Lower bound verified}: For each of the 16 F16 operators, we exhibit a
  concrete witness $(x_0, y_0)$ such that $\text{op}(x_0, y_0) \neq f(x_0, y_0)$
  (where $f$ is the target function). Each witness proof is 1-2 tactic lines.
  All 16 witnesses per operation compile with 0 sorries.

\item \textbf{Upper bound verified}: Explicit algebraic identity, proved by
  Lean's \texttt{rw} + \texttt{simp} + \texttt{ring\_nf} tactics using Mathlib
  lemmas for \texttt{Real.exp}, \texttt{Real.log}, \texttt{Real.sqrt}, \texttt{Real.rpow}.

\item \textbf{Build status}: \texttt{lake build → Build completed successfully} (2026-04-21).
\end{itemize}

\section{Correcting the v5.1 Overcounting}

\begin{theorem}[v5.1 overcounting, \lv{ModelAudit.lean}]
The SuperBEST v5.1 positive-domain total of 16n is overcounted by 1.
sqrt($x$) = EPL(0.5, $x$) = $\exp(0.5\ln x)$ is a single F16 node for $x > 0$.
Corrected positive total: $15$n (savings: 79.5\%).
\end{theorem}

\noindent This correction is now Lean-verified (\texttt{superbest\_v51\_overcounted\_by\_one}).
The v5.2 table above reflects the corrected count.

\section{The 23-Operator Extended Set — Convenience Layer Only}

\textbf{Policy (from CONTEXT.md, effective immediately):}

\begin{itemize}
\item The 23-operator set (F16 $\cup$ \{EEM, EED, EES, EEA\} $\cup$ \{LLA, LLS, LLD\})
  is a \emph{Layer 2 / convenience layer} for library design, ML papers, and
  efficiency comparisons.
\item \emph{No formal lower bound theorem} uses the 23-op count. All formal
  proofs are stated over F16 (16 operators).
\item When using 23-op costs, tables must be labeled ``23-op count'' or
  ``extended-primitive count'' explicitly. Never mix F16 and 23-op counts in
  the same unlabeled column.
\item The 7 extended operators (EEM, EED, EES, EEA, LLA, LLS, LLD) count as
  1n only if atomically implemented. Their F16 costs are:
  EEM=3n, EED=3n, EES=1n (special case $y=0$)/4n (general),
  LLA=3n, LLS=3n, LLD=4n, EEA=4n.
\end{itemize}

\section{Gap: Tight Mul Lower Bound (CONJ\_MUL\_GEN\_TIGHT)}

The Lean-verified lower bound is SB(mul, general) $\geq$ 2.
The known upper bound is SB(mul, general) $\leq$ 6.
The exact cost in $[2, 6]$ is open.

\begin{itemize}
\item \textbf{Known}: $\geq 2$ (Lean: MulLowerBound.lean, SB\_mul\_ge\_two, 0 sorries)
\item \textbf{Known}: $\leq 6$ (upper: neg(mul(neg($x$),$y$)) = $xy$, 6 nodes)
\item \textbf{Open}: Is the exact cost 6? Or can it be done in 3, 4, or 5 nodes?
\item \textbf{Immediate subgoal}: Prove $\geq 3$ by extending the witness approach
  to 2-node trees (there are $|F16|^2 = 256$ compositions to check; likely
  automatable with a small Lean tactic or Python exhaustive search).
\end{itemize}

\section{Recommended Next Step: Automated Lower Bound Checker}

The current witness proofs are manual (one theorem per operator). A natural automation:

\begin{enumerate}
\item Write a Lean tactic \texttt{decide\_lower\_bound} that, given target $f$
  and bound $k$, enumerates all $k$-node F16 trees and checks each against
  a finite set of test points.
\item Use \texttt{native\_decide} for small $k$ (tree count is $O(|F16|^k)$).
\item This would automatically verify $k$-node lower bounds for mul, div, and
  any other target, closing CONJ\_MUL\_GEN\_TIGHT.
\end{enumerate}

\end{document}
